---
title: "Host_Genetics_Analysis"
---


#Make sure you have all the files needed to start. If you using PLINK files. Make sure you have the .bim, .fam and .bed files.

#Also have the reference to which the genome was mapped. Remember to index the refence before use

Examples

`{=} H3_Africa_order_3220_original_annotation.anno.txt  H3_Africa_order_3220_original_annotation.fam H3_Africa_order_3220_original_annotation.bed        H3_Africa_order_3220_original_annotation.bim        human_g1k_v37.fasta human_g1k_v37.fasta.fai` \## Quick sanity check: build & contigs

Just to be sure everything matches:

```{bash}
# Check chromosome naming in your BIM
head H3_Africa_order_3220_original_annotation.bim

# Check contigs in the reference FASTA index
head human_g1k_v37.fasta.fai
```

## 1. Remove ambiguous strand SNPs (A/T and C/G)

These SNPs are impossible to orient confidently (A↔T, C↔G look the same on the opposite strand), so best practice is to **remove them before strand alignment**.

```{bash}
# 1. Make a list of ambiguous SNPs (A/T and C/G) from the BIM
awk '($5$6=="AT" || $5$6=="TA" || $5$6=="CG" || $5$6=="GC") {print $2}' \
    H3_Africa_order_3220_original_annotation.bim > ambiguous_AT_CG.snps

wc -l ambiguous_AT_CG.snps   # how many are ambiguous

# 2. Create a cleaned PLINK set without these problematic SNPs
plink --bfile H3_Africa_order_3220_original_annotation \
      --exclude ambiguous_AT_CG.snps \
      --make-bed \
      --out H3_noAmbig
```

Now `H3_noAmbig.{bed,bim,fam}` is your **working dataset** without A/T or C/G SNPs.

**Best practices : Always have a QC_tracker**

This will help ypu update a CSV file after each major step and records items such as variants remaining, variants excluded, number of individuals (males vs females) number of cases and controls etc.

You can modify and run the below R script after doing these 2 after each step

```{bash}
plink --bfile H3_noAmbig --missing --out H3_noAmbig

Rscript qc_tracker.R H3_noAmbig "Removed ambiguous A/T & C/G SNPs"
```

Save the below Rscript as **qc.tracker.R**

```{r}
#!/usr/bin/env Rscript

############################################################
# QC PROGRESS TRACKER (with filter metadata)
# ------------------------------------------
# Usage:
#   Rscript qc_tracker.R <plink_prefix> "<step_name>" [qc_csv] [options]
#
# Options (all optional, free text allowed):
#   --snp-miss <value>       # e.g. 0.05 for --geno 0.05
#   --sample-miss <value>    # e.g. 0.05 for --mind 0.05
#   --maf <value>            # e.g. 0.01 for --maf 0.01
#   --hwe <value>            # e.g. 1e-6 for --hwe 1e-6
#   --info <value>           # e.g. 0.8 for post-imputation info filter
#   --other "<free text>"    # any notes, multiple words allowed
#
# Examples:
#   Rscript qc_tracker.R H3_noAmbig "Removed ambiguous SNPs"
#
#   Rscript qc_tracker.R H3_mind0.05 "Sample missingness filter" QC_progress_tracker.csv \
#       --sample-miss 0.05
#
#   Rscript qc_tracker.R H3_maf0.01_hwe1e-6 "MAF + HWE filter" QC_progress_tracker.csv \
#       --maf 0.01 --hwe 1e-6 --snp-miss 0.05 \
#       --other "Autosomes only; removed monomorphic SNPs"
#
############################################################



## 2. Convert to VCF and align alleles to human_g1k_v37.fasta

Here we use **bcftools +fixref** to ensure the **REF allele** always matches the base in `human_g1k_v37.fasta`.\
If the variant is on the opposite strand, `fixref` will complement alleles; if REF/ALT are swapped, it will fix that too.

### **2a. PLINK → VCF**

```{bash}
plink --bfile H3_noAmbig \
      --recode vcf bgz \
      --out H3_noAmbig_raw

# creates H3_noAmbig_raw.vcf.gz
```

### 2b. Use bcftools +fixref to fix strand & REF mismatches

```{bash}
# Index the VCF (bcftools likes it indexed)
bcftools index H3_noAmbig_raw.vcf.gz

# Run fixref:
#  -f human_g1k_v37.fasta : reference genome
#  -m top : try to flip strand and/or swap REF/ALT so that REF matches fasta
#  -d     : drop variants that still cannot be fixed
bcftools +fixref H3_noAmbig_raw.vcf.gz \
    -Oz -o H3_refAligned.vcf.gz \
    -- -f human_g1k_v37.fasta -m top -d

# Index the fixed VCF
bcftools index H3_refAligned.vcf.gz
```

What this does conceptually:

-   For every variant, it looks at `CHR:POS` in the FASTA.

-   If your REF is wrong but its **complement** matches the FASTA, it **flips strand**.

-   If REF/ALT are swapped, it swaps them.

-   If it still can’t make your REF match the reference base, it **drops** the variant

-   You can see how many were dropped by:

```{bash}
bcftools view -H H3_noAmbig_raw.vcf.gz | wc -l
bcftools view -H H3_refAligned.vcf.gz  | wc -l
```

Difference = number of unresolved mismatches removed.

In this H3_noAmbig_raw = 2186579\
H3_refAlignd= 2141373

Note: After conversion to VCF and conversion back to PLINK, phenotype data (case/control) and sex (male/female) are lost. You will need to copy them from your original fam file into the new one created.

### 2c. Rscript to restore sex & phenotype

Save this as `restore_fam_metadata.R` in your working directory:\
Note: FID and IID in new .fam may differ from the old. Example old starts with 2...... where a new may start with 0_2..... The below scripts corects it to conform with the initial IDs

```{r}
#!/usr/bin/env Rscript

############################################################
# restore_fam_metadata.R
# ----------------------
# For your dataset:
#   - Original FAM: H3_Africa_order_3220_original_annotation.fam
#   - New FAM     : H3_GRCh37_refAligned.fam
#
# Does:
#   1) Copies SEX and PHENO from original to new *by row position*
#   2) Strips leading "0_" from FID and IID (0_2503... -> 2503...)
#   3) Backs up new FAM → H3_GRCh37_refAligned.fam.bak
#   4) Overwrites H3_GRCh37_refAligned.fam with corrected values
############################################################

suppressPackageStartupMessages({
  library(data.table)
})

orig_fam_path <- "H3_Africa_order_3220_original_annotation.fam"
new_fam_path  <- "H3_GRCh37_refAligned.fam"

cat("Original FAM:", orig_fam_path, "\n")
cat("New FAM     :", new_fam_path, "\n\n")

if (!file.exists(orig_fam_path)) stop("Original FAM not found: ", orig_fam_path)
if (!file.exists(new_fam_path))  stop("New FAM not found: ", new_fam_path)

# Read both FAMs as character
orig <- fread(
  orig_fam_path,
  header     = FALSE,
  colClasses = rep("character", 6),
  col.names  = c("FID", "IID", "PAT", "MAT", "SEX", "PHENO")
)

new <- fread(
  new_fam_path,
  header     = FALSE,
  colClasses = rep("character", 6),
  col.names  = c("FID", "IID", "PAT", "MAT", "SEX", "PHENO")
)

cat("Original FAM samples:", nrow(orig), "\n")
cat("New FAM samples     :", nrow(new), "\n\n")

if (nrow(orig) != nrow(new)) {
  stop("Number of samples differ between original and new FAM - cannot copy by position.")
}

# 1) Copy SEX and PHENO by row position
new$SEX   <- orig$SEX
new$PHENO <- orig$PHENO

# 2) Strip leading "0_" from FID and IID, if present
new$FID <- sub("^0_", "", new$FID)
new$IID <- sub("^0_", "", new$IID)

cat("Example rows after copying + ID fix:\n")
print(new[1:5, ])

# Backup old FAM
backup_path <- paste0(new_fam_path, ".bak")
file.copy(new_fam_path, backup_path, overwrite = TRUE)
cat("\nBackup written to:", backup_path, "\n")

# Overwrite the new FAM with corrected metadata
fwrite(
  new,
  file      = new_fam_path,
  sep       = " ",
  col.names = FALSE,
  quote     = FALSE
)

cat("✅ Corrected FAM written to:", new_fam_path, "\n")
cat("   - SEX and PHENO copied from original (by row)\n")
cat("   - FID/IID cleaned (removed '0_' prefix)\n")
```

-   FID / IID = `2503…` (no `0_`)

-   SEX = 1/2 (male/female)

-   PHENO = 1/2 (control/case)

Once that looks good:

Record all the steps into the qc_tracker.csv

```{bash}
plink --bfile H3_GRCh37_refAligned \
      --missing \
      --out H3_GRCh37_refAligned

Rscript qc_tracker.R H3_GRCh37_refAligned \
    "Post-strand alignment + restored sex/phenotype + fixed IDs"
```

Great, now we have our perfectly aligned files. We can move on to Quality Control steps.

# 3. Quality Control Steps

## Sample Level Quality Control

First run a baseline snapshot of your data beofre QC. And log into your qc_tracker

```{bash}
plink --bfile H3_GRCh37_refAligned \
      --missing \
      --out H3_GRCh37_refAligned

Rscript qc_tracker.R H3_GRCh37_refAligned \
    "Baseline: post-strand alignment + fixed IDs/phenotypes"
```

## SNP missingness (first `–geno 0.05`)

Remove SNPs missing in \>5% of samples

```{bash}
plink --bfile H3_GRCh37_refAligned \
      --geno 0.05 \
      --make-bed \
      --out H3_QC1_geno0.05

plink --bfile H3_QC1_geno0.05 \
      --missing \
      --out H3_QC1_geno0.05

Rscript qc_tracker.R H3_QC1_geno0.05 \
    "Step 1: SNP call-rate filter (--geno 0.05)" \
    QC_progress_tracker.csv \
    --snp-miss 0.05
```

## Sample missingness (`--mind 0.05`)

Remove individuals with \>5% missing genotypes.

```{bash}
plink --bfile H3_QC1_geno0.05 \
      --mind 0.05 \
      --make-bed \
      --out H3_QC2_mind0.05

plink --bfile H3_QC2_mind0.05 \
      --missing \
      --out H3_QC2_mind0.05

Rscript qc_tracker.R H3_QC2_mind0.05 \
    "Step 2: Sample call-rate filter (--mind 0.05)" \
    QC_progress_tracker.csv \
    --sample-miss 0.05
```

## SNP missingness again (second `--geno 0.05`)

Now that bad individuals are gone, SNP missingness is recalculated; some SNPs will newly cross 5% missing and should be dropped.

```{bash}
plink --bfile H3_QC2_mind0.05 \
      --geno 0.05 \
      --make-bed \
      --out H3_QC3_geno0.05_again

plink --bfile H3_QC3_geno0.05_again \
      --missing \
      --out H3_QC3_geno0.05_again

Rscript qc_tracker.R H3_QC3_geno0.05_again \
    "Step 3: SNP call-rate filter again (--geno 0.05) after sample cleanup" \
    QC_progress_tracker.csv \
    --snp-miss 0.05
```

## MAF filter (`--maf 0.01`)

Keep variants with minor allele frequency ≥1%.

```{bash}
plink --bfile H3_QC3_geno0.05_again \
      --maf 0.01 \
      --make-bed \
      --out H3_QC4_maf0.01

plink --bfile H3_QC4_maf0.01 \
      --missing \
      --out H3_QC4_maf0.01

Rscript qc_tracker.R H3_QC4_maf0.01 \
    "Step 4: MAF filter (--maf 0.01)" \
    QC_progress_tracker.csv \
    --maf 0.01
```

## HWE filter (`--hwe 1e-6 midp` in controls)

Only done in controls

```{bash}
plink --bfile H3_QC4_maf0.01 \
      --hwe 1e-6 midp \
      --make-bed \
      --out H3_QC5_hwe1e-6

plink --
bfile H3_QC5_hwe1e-6 \
      --missing \
      --out H3_QC5_hwe1e-6

Rscript qc_tracker.R H3_QC5_hwe1e-6 \
    "Step 5: HWE filter in controls (--hwe 1e-6 midp)" \
    QC_progress_tracker.csv \
    --hwe 1e-6 \
    --other "HWE midp; applied to controls only"
```

## Autosomes only (chr 1–22)

For PCA, imputation, GWAS keep only autosomes.

```{bash}
plink --bfile H3_QC5_hwe1e-6 \
      --chr 1-22 \
      --make-bed \
      --out H3_QC6_autosomes

plink --bfile H3_QC6_autosomes \
      --missing \
      --out H3_QC6_autosomes

Rscript qc_tracker.R H3_QC6_autosomes \
    "Step 6: Restricted to autosomes (chr 1–22)" \
    QC_progress_tracker.csv \
    --other "chr 1–22 only"
```

## **Relatedness (IBD) and removal**

```{bash}
plink --bfile H3_QC6_autosomes \
      --het \
      --out H3_QC7_het
```

### Identify outliers in R (\|F z-score\| \> 3)

Create a small R script, e.g. `find_het_outliers.R`:

```{r}
#!/usr/bin/env Rscript

suppressPackageStartupMessages({
  library(data.table)
})

het <- fread("H3_QC7_het.het")

# F = inbreeding coefficient; use z-score on F
het[, mean_F := mean(F, na.rm = TRUE)]
het[, sd_F   := sd(F, na.rm = TRUE)]
het[, z_F    := (F - mean_F) / sd_F]

outliers <- het[abs(z_F) > 3, .(FID, IID, F, z_F)]

cat("Total samples:", nrow(het), "\n")
cat("Heterozygosity outliers (|z_F| > 3):", nrow(outliers), "\n")

if (nrow(outliers) > 0) {
  fwrite(outliers[, .(FID, IID)],
         file = "het_fail_remove.txt",
         sep = "\t",
         col.names = FALSE)
  cat("Wrote outliers to het_fail_remove.txt\n")
} else {
  cat("No heterozygosity outliers detected. No file written.\n")
}
```

### Remove outliers (if any) and track

If `het_fail_remove.txt` **exists and is non-empty**:

```{bash}
plink --bfile H3_QC6_autosomes \
      --remove het_fail_remove.txt \
      --make-bed \
      --out H3_QC7_hetClean

plink --bfile H3_QC7_hetClean \
      --missing \
      --out H3_QC7_hetClean

Rscript qc_tracker.R H3_QC7_hetClean \
    "Step 7: Removed heterozygosity outliers (|F z| > 3)" \
    QC_progress_tracker.csv \
    --other "Removed samples with |F z-score| > 3"
```

## Relatedness (IBD) and removal of relatives

### Compute IBD

```{bash}
plink --bfile H3_QC7_hetClean \
      --genome \
      --min 0.185 \
      --out H3_QC8_ibd
```

This makes `H3_QC8_ibd.genome` with pairwise IBD estimates.

### Select one to remove per related pair in R

R script `find_ibd_related.R`:

```{r}
#!/usr/bin/env Rscript

suppressPackageStartupMessages({
  library(data.table)
})

ibd <- fread("H3_QC8_ibd.genome")

# PI_HAT > 0.185 ~ 2nd-degree or closer
pairs <- ibd[PI_HAT > 0.185]

cat("Total pairs with PI_HAT > 0.185:", nrow(pairs), "\n")

if (nrow(pairs) == 0) {
  cat("No related pairs found. No file written.\n")
  quit(save = "no")
}

# Simple heuristic: remove the second individual in each pair
to_remove <- unique(pairs[, .(FID = FID2, IID = IID2)])

cat("Individuals to remove:", nrow(to_remove), "\n")

fwrite(to_remove,
       file = "ibd_fail_remove.txt",
       sep = "\t",
       col.names = FALSE)

cat("Wrote related individuals to ibd_fail_remove.txt\n")
```

Input a final track into `QC_progress_tracker.csv`

```{bash}
Rscript qc_tracker.R H3_QC8_ibdClean \
    "Step 9: Final QC-clean dataset (pre-imputation / pre-GWAS)" \
    QC_progress_tracker.csv \
    --snp-miss 0.05 --sample-miss 0.05 --maf 0.01 --hwe 1e-6 \
    --other "Autosomes only; post heterozygosity and IBD filters"
```

# 4. Population Structure Determination (PCA & Admixture)

## Setup

We create indivdual directories for PCA and Admixture to keep things clean.

From the directory with `H3_QC8_ibdClean.*`:

```{bash}
mkdir -p PCA ADMIXTURE
```

Note: If you have a differenet name for your post QC files. Replace H3_QC8_ibdClean with your file name for the following steps.

## PCA PIPELINE (into `PCA/`)

### LD pruning for PCA

```{bash}
plink --bfile H3_QC8_ibdClean \
      --indep-pairwise 50 5 0.2 \
      --out PCA/H3_QC8_prune

plink --bfile H3_QC8_ibdClean \
      --extract PCA/H3_QC8_prune.prune.in \
      --make-bed \
      --out PCA/H3_QC8_pruned
```

### Run PCA with PLINK

```{bash}
plink --bfile PCA/H3_QC8_pruned \
      --pca 20 \
      --out PCA/H3_PCA
```

Don't forget to login the changes into the qc_tracker

```{bash}
Rscript qc_tracker.R PCA/H3_QC8_pruned \
    "Step 11: PCA computed on LD-pruned SNPs" \
    QC_progress_tracker.csv \
    --other "PCA: 20 components; PLINK exact PCA; LD pruning: 50 5 0.2; PCA output prefix: PCA/H3_PCA"
```

Now make sure you have your phenotype data in .txt or .tsv format. It can contain things like ethnicity, age, strain_type etc.

For this we used a phenotype data called H3_pheno.phe.txt which contained the above parameters. We will then merge with phenotype data and plot the PCA in R as below.

```{r}
#!/usr/bin/env Rscript

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(cowplot)  # for combining A/B panels
})

# -------------------------------------------------------
# 1) Read PCA eigenvectors
# -------------------------------------------------------
eig <- fread("PCA/H3_PCA.eigenvec")
# PLINK format: FID IID PC1 PC2 PC3 ...
colnames(eig) <- c("FID", "IID", paste0("PC", 1:(ncol(eig) - 2)))

cat("Read PCA eigenvec with", nrow(eig), "samples and",
    ncol(eig) - 2, "PCs.\n")

# -------------------------------------------------------
# 2) Read eigenvalues and compute variance explained
# -------------------------------------------------------
eigenval <- fread("PCA/H3_PCA.eigenval", header = FALSE)$V1
variance_explained <- eigenval / sum(eigenval)

pc1_var <- round(variance_explained[1] * 100, 2)
pc2_var <- round(variance_explained[2] * 100, 2)

pc1_lab <- paste0("PC1 (", pc1_var, "% variance)")
pc2_lab <- paste0("PC2 (", pc2_var, "% variance)")

cat("PC1 variance:", pc1_var, "%\n")
cat("PC2 variance:", pc2_var, "%\n")

# -------------------------------------------------------
# 3) Read phenotype file
#    H3_pheno.phe.txt with columns:
#    FID IID AGE REGION ETHNICITY STRAIN
# -------------------------------------------------------
pheno <- fread("PCA/H3_pheno.phe.txt")

cat("Read phenotype file with", nrow(pheno), "rows.\n")
cat("Phenotype columns:\n")
print(colnames(pheno))

# -------------------------------------------------------
# 4) Merge PCA + phenotype
# -------------------------------------------------------
dat <- eig %>%
  left_join(pheno, by = c("FID", "IID"))

cat("Merged PCA + pheno; rows:", nrow(dat), "\n\n")

# -------------------------------------------------------
# 5) Recode ETHNICITY: Akan, Ewe, Ga, others -> "Other"
# -------------------------------------------------------
dat <- dat %>%
  mutate(
    ETHNICITY_CLEAN = case_when(
      !is.na(ETHNICITY) & tolower(ETHNICITY) == "akan" ~ "Akan",
      !is.na(ETHNICITY) & tolower(ETHNICITY) == "ewe"  ~ "Ewe",
      !is.na(ETHNICITY) & tolower(ETHNICITY) == "ga"   ~ "Ga",
      is.na(ETHNICITY) ~ NA_character_,
      TRUE ~ "Other"
    ),
    ETHNICITY_CLEAN = factor(
      ETHNICITY_CLEAN,
      levels = c("Akan", "Ewe", "Ga", "Other")
    )
  )

# -------------------------------------------------------
# 6) Recode STRAIN: Control, MAF, Mtbss
# -------------------------------------------------------
dat <- dat %>%
  mutate(
    STRAIN_CLEAN = case_when(
      !is.na(STRAIN) & grepl("^control$", STRAIN, ignore.case = TRUE) ~ "Control",
      !is.na(STRAIN) & grepl("maf", STRAIN, ignore.case = TRUE)       ~ "MAF",
      !is.na(STRAIN) & grepl("mtb", STRAIN, ignore.case = TRUE)       ~ "Mtbss",
      TRUE ~ STRAIN
    ),
    STRAIN_CLEAN = factor(
      STRAIN_CLEAN,
      levels = c("Control", "MAF", "Mtbss")
    )
  )

# -------------------------------------------------------
# 7) PCA plot A: ETHNICITY
# -------------------------------------------------------
p_eth <- dat %>%
  filter(!is.na(ETHNICITY_CLEAN)) %>%
  ggplot(aes(x = PC1, y = PC2, color = ETHNICITY_CLEAN)) +
  geom_point(alpha = 0.9, size = 1) +
  labs(
    title = "PCA coloured by Ethnicity",
    x = pc1_lab,
    y = pc2_lab,
    color = "Ethnicity"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_line(size = 0.2),
    legend.position = "right"
  )

ggsave("PCA/H3_PCA_PC1_PC2_ETHNICITY.png",
       p_eth, width = 7, height = 5, dpi = 300)

cat("Saved PCA ethnicity plot to PCA/H3_PCA_PC1_PC2_ETHNICITY.png\n")

# -------------------------------------------------------
# 8) PCA plot B: STRAIN
# -------------------------------------------------------
p_strain <- dat %>%
  filter(!is.na(STRAIN_CLEAN)) %>%
  ggplot(aes(x = PC1, y = PC2, color = STRAIN_CLEAN)) +
  geom_point(alpha = 0.9, size = 1) +
  labs(
    title = "PCA coloured by Strain",
    x = pc1_lab,
    y = pc2_lab,
    color = "Strain"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_line(size = 0.2),
    legend.position = "center"
  )

ggsave("PCA/H3_PCA_PC1_PC2_STRAIN.png",
       p_strain, width = 7, height = 5, dpi = 300)

cat("Saved PCA strain plot to PCA/H3_PCA_PC1_PC2_STRAIN.png\n")

# -------------------------------------------------------
# 9) Combined panel (A = ethnicity, B = strain)
# -------------------------------------------------------
combined <- plot_grid(
  p_eth, p_strain,
  ncol = 1,
  labels = c("A", "B"),
  label_size = 14,
  label_fontface = "bold",
  rel_heights = c(1, 1)
)

ggsave("PCA/H3_PCA_PC1_PC2_ETHNICITY_STRAIN_combined.png",
       combined, width = 7, height = 9, dpi = 300)

cat("Saved combined PCA panel to PCA/H3_PCA_PC1_PC2_ETHNICITY_STRAIN_combined.png\n")
```

### Merging our DATA with that of the 1000 Genomes Project

```{bash}
plink --bfile H3_QC8_ibdClean \
      --bmerge PCA/1000G_merged.bed PCA/1000G_merged.bim PCA/1000G_merged.fam \
      --make-bed \
      --out PCA/Ghana_1000G_tmp
```

You can get low genotyping rate as many of the SNPs found in the 1000 genomes project may not be found in our dataset

Restrict to **autosomes** and **geno 0.05** to effectively keep SNPs shared and well-called in both cohorts.

```{bash}
plink --bfile PCA/Ghana_1000G_tmp \
      --chr 1-22 \
      --geno 0.05 \
      --make-bed \
      --out PCA/Ghana_1000G_geno0.05_autosomes
```

An error occurs where all samples are removed bt produces a file

After map the chromosome positions in each file better

```{bash}
plink --bfile H3_QC8_ibdClean_posID \ 
      --bmerge PCA/1000G_merged_posID.bed PCA/1000G_merged_posID.bim PCA/1000G_merged_posID.fam \ 
      --make-bed \ 
      --out PCA/Ghana_1000G_posID_tmp
```

Remove multiallelic variants

```{bash}
plink --bfile PCA/1000G_merged_posID
      --exclude PCA/Ghana_1000G_posID_tmp-merge.missnp
      --make-bed
      --out PCA/1000G_merged_posID_nomulti
```

```{bash}
plink --bfile H3_QC8_ibdClean_posID_nomulti
      --bmerge PCA/1000G_merged_posID_nomulti.bed PCA/1000G_merged_posID_nomulti.bim PCA/1000G_merged_posID_nomulti.fam
      --make-bed
       --out PCA/Ghana_1000G_posID_merged
```

Prune data before PCA

```{bash}
plink --bfile PCA/Ghana_1000G_posID_merged
      --indep-pairwise 50 5 0.2
      --out PCA/Ghana_1000G_prune
```

Extract pruned data to make bed, bim and fam files

```{bash}
plink --bfile PCA/Ghana_1000G_posID_merged
      --extract PCA/Ghana_1000G_prune.prune.in
      --make-bed
      --out PCA/Ghana_1000G_pruned
```

Run PCA on pruned data

```{bash}
plink --bfile PCA/Ghana_1000G_pruned
      --out PCA/Ghana_1000G_PCA
      --pca 20
```

R script to plot merged 1000 genomes and Personal data

```{r}
#!/usr/bin/env Rscript

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
})

# -------------------------------------------------------
# 1) Read PCA eigenvectors & eigenvalues
# -------------------------------------------------------
eig <- fread("PCA/Ghana_1000G_PCA.eigenvec")
colnames(eig) <- c("FID", "IID", paste0("PC", 1:(ncol(eig) - 2)))

# Ensure IDs are character
eig <- eig %>%
  mutate(
    FID = as.character(FID),
    IID = as.character(IID)
  )

cat("Read merged PCA eigenvec with", nrow(eig), "samples and",
    ncol(eig) - 2, "PCs.\n")

eigenval <- fread("PCA/Ghana_1000G_PCA.eigenval", header = FALSE)$V1
variance_explained <- eigenval / sum(eigenval)

pc1_var <- round(variance_explained[1] * 100, 2)
pc2_var <- round(variance_explained[2] * 100, 2)

pc1_lab <- paste0("PC1 (", pc1_var, "% variance)")
pc2_lab <- paste0("PC2 (", pc2_var, "% variance)")

cat("PC1 variance:", pc1_var, "%\n")
cat("PC2 variance:", pc2_var, "%\n")

# -------------------------------------------------------
# 2) Ghana phenotype file (marks Ghana samples)
#    H3_pheno.phe.txt: FID IID AGE REGION ETHNICITY STRAIN
# -------------------------------------------------------
ghana_pheno <- fread("PCA/H3_pheno.phe.txt")

ghana_pheno <- ghana_pheno %>%
  mutate(
    FID = as.character(FID),
    IID = as.character(IID)
  )

ghana_ids <- ghana_pheno %>%
  select(FID, IID) %>%
  mutate(GROUP_GHANA = "Ghana")

cat("Ghana samples in pheno:", nrow(ghana_ids), "\n")

# -------------------------------------------------------
# 3) 1000 Genomes datasheet
#    Columns: sample, pop, super_pop, gender
# -------------------------------------------------------
kg <- fread("PCA/1000_Genomes_Datasheet.txt")

cat("1000 Genomes datasheet rows:", nrow(kg), "\n")
cat("Datasheet columns:\n")
print(colnames(kg))

# Build IID + subpopulation annotation (as character)
kg_annot <- kg %>%
  transmute(
    IID       = as.character(sample),
    KG_SUBPOP = super_pop
  )

# -------------------------------------------------------
# 4) Merge PCA with Ghana + 1000G annotations
# -------------------------------------------------------
dat <- eig %>%
  # mark Ghana samples (FID+IID)
  left_join(ghana_ids, by = c("FID", "IID")) %>%
  # add 1000G subpop by IID
  left_join(kg_annot, by = "IID") %>%
  mutate(
    GROUP = case_when(
      !is.na(GROUP_GHANA) ~ "Ghana",
      !is.na(KG_SUBPOP)   ~ KG_SUBPOP,
      TRUE                ~ "Unknown"
    )
  )

cat("Merged PCA + annotations; rows:", nrow(dat), "\n")
cat("Group counts:\n")
print(table(dat$GROUP))

# -------------------------------------------------------
# 5) PCA plot: Ghana + 1000G subpopulations
# -------------------------------------------------------
p <- dat %>%
  filter(GROUP != "Unknown") %>%
  ggplot(aes(x = PC1, y = PC2, color = GROUP)) +
  geom_point(alpha = 0.9, size = 1.0) +
  labs(
    title = "PCA: Ghana + 1000 Genomes subpopulations",
    x = pc1_lab,
    y = pc2_lab,
    color = "Group"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_line(size = 0.2),
    legend.position = "right"
  )

# Make sure output dir exists (PCA/)
if (!dir.exists("PCA")) dir.create("PCA", recursive = TRUE)

ggsave("PCA/PCA_1000G_merge/Ghana_1000G_PCA_PC1_PC2_GROUP.png",
       p, width = 8, height = 6, dpi = 300)

cat("Saved PCA plot to PCA/PCA_1000G_merge/Ghana_1000G_PCA_PC1_PC2_GROUP.png\n")
```

Log in all your PCA steps into `QC_progress_tracker.csv`

```{bash}
# 1) Ghana posID after removing multiallelic variants
plink --bfile H3_QC8_ibdClean_posID_nomulti \
      --missing \
      --out H3_QC8_ibdClean_posID_nomulti

# 2) 1000G posID after removing multiallelic variants
plink --bfile PCA/1000G_merged_posID_nomulti \
      --missing \
      --out PCA/1000G_merged_posID_nomulti

# 3) Final merged Ghana+1000G dataset
plink --bfile PCA/Ghana_1000G_posID_merged \
      --missing \
      --out PCA/Ghana_1000G_posID_merged

# 4) LD-pruned final dataset (used for PCA/ADMIXTURE)
plink --bfile PCA/Ghana_1000G_pruned \
      --missing \
      --out PCA/Ghana_1000G_pruned
```

#### Log: removal of problematic SNPs (Ghana side)

**Dataset:** `H3_QC8_ibdClean_posID_nomulti`

```{bash}
Rscript qc_tracker.R H3_QC8_ibdClean_posID_nomulti \
  "Step 12: Ghana posID dataset after removing multiallelic / discordant SNPs for merge" \
  QC_progress_tracker.csv \
  --other "Started from H3_QC8_ibdClean_posID; excluded SNPs in PCA/Ghana_1000G_posID_tmp-merge.missnp"
```

#### Log: removal of problematic SNPs (1000G side)

**Dataset:** `PCA/1000G_merged_posID_nomulti`

```{bash}
Rscript qc_tracker.R PCA/1000G_merged_posID_nomulti \
  "Step 13: 1000 Genomes posID dataset after removing multiallelic / discordant SNPs for merge" \
  QC_progress_tracker.csv \
  --other "Started from PCA/1000G_merged_posID; excluded SNPs in PCA/Ghana_1000G_posID_tmp-merge.missnp"
```

#### Log: final merged Ghana + 1000G dataset (before pruning)

**Dataset:** `PCA/Ghana_1000G_posID_merged`

This is the key PCA base dataset.

```{bash}
Rscript qc_tracker.R PCA/Ghana_1000G_posID_merged \
  "Step 14: Final merged Ghana + 1000G dataset (posID, GRCh37)" \
  QC_progress_tracker.csv \
  --other "Successfully merged H3_QC8_ibdClean_posID_nomulti with 1000G_merged_posID_nomulti; includes Ghana + 2504 1000G samples"
```

#### Log: LD-pruned set used for PCA & ADMIXTURE

**Dataset:** `PCA/Ghana_1000G_pruned`

```{bash}
Rscript qc_tracker.R PCA/Ghana_1000G_pruned \
  "Step 15: LD-pruned Ghana + 1000G dataset for PCA/ADMIXTURE" \
  QC_progress_tracker.csv \
  --other "indep-pairwise 50 5 0.2 on PCA/Ghana_1000G_posID_merged; this SNP set used for PCA and ADMIXTURE"
```

#### Log: PCA step itself (on pruned dataset)

PCA doesn’t change SNPs/samples, but we want the parameters in the log. We point qc_tracker again to the **pruned PLINK prefix** and just change the description + `--other`.

```{bash}
Rscript qc_tracker.R PCA/Ghana_1000G_pruned \
  "Step 16: PCA on Ghana + 1000G LD-pruned dataset" \
  QC_progress_tracker.csv \
  --other "PCA: 20 components from PCA/Ghana_1000G_pruned; outputs=Ghana_1000G_PCA.eigenvec/eigenval; variance PC1≈$57.67%, PC2≈$19.9%, PC3≈$6.09%"
```

#### (Optional but nice) Log: Ghana-only PCA with ethnicity

Even though this uses the **same eigenvec file**, it’s conceptually another analysis step: “PCA interpreted within Ghana; coloured by ETHNICITY”. If you want that documented:

```{bash}
Rscript qc_tracker.R PCA/Ghana_1000G_pruned \
  "Step 17: Ghana-only PCA interpretation with ethnicity (Akan/Ewe/Ga/Other)" \
  QC_progress_tracker.csv \
  --other "Used Ghana subset of Ghana_1000G_PCA eigenvectors; ETHNICITY from H3_pheno.phe.txt; produced ethnicity-stratified PCA plots"
```

## ADMIXTURE Analysis

You can run the admixture on the merged H3_Africa and 1000 Genomes sample `Ghana_1000G_pruned.bed`

```{bash}
for K in {2..10}; do
  echo ">>> Running ADMIXTURE for K=${K}"
  admixture --cv=10 Ghana_1000G_pruned.bed $K | tee admixture_K${K}.log
done
```

You can also run admixture on just your pruned SNPs without 1000 genomes merge:

```{bash}
for K in {2..10}; do
  echo ">>> Running ADMIXTURE for K=${K}"
  admixture --cv=10 Ghana_1000G_pruned.bed $K | tee admixture_K${K}.log
done
```

# 5. Preparation for Imputation

## Liftover GRCh37 → GRCh38 using CrossMap

Use your **clean, pre-pruned GRCh37 dataset:** `H3_QC8_ibdClean.{bed,bim,fam}`

Create new a directory and copy `H3_QC8_ibdClean.{bed,bim,fam}` into this directory. Start working from the directory containing these files.

```{bash}
mkdir -p Liftover_GRCh38
cd Liftover_GRCh38

cd ..
cp H3_QC8_ibdClean.b* Liftover_GRCh38/
cp H3_QC8_ibdClean.fam Liftover_GRCh38/

cd Liftover_GRCh38
```

Now we have `H3_QC8_ibdClean.{bed,bim,fam}` in the Liftover_GRCh38 directory

## Convert PLINK (GRCh37) → VCF

Use PLINK to export a VCF with your current genotypes and index it:

```{bash}
plink --bfile H3_QC8_ibdClean \
      --recode vcf bgz \
      --out H3_QC8_ibdClean_GRCh37
      
tabix -p vcf H3_QC8_ibdClean_GRCh37.vcf.gz
```

It produces two files `H3_QC8_ibdClean.GRCh37.vcf.gz` and `H3_QC8_ibdClean_GRCh37.vcf.gz.tbi`

Create a conda environment for crossmap. This is used for ...

```{bash}
conda create -n crossmap_env -c bioconda -c conda-forge python=3.10 crossmap

conda activate crossmap_env
```

### Downloading the GRCh38 reference genome from Ensemble

```{bash}
mkdir -p /home/tafum/references/GRCh38
cd /home/tafum/references/GRCh38
```

```{bash}
# Ensembl-style primary assembly (example URL pattern)
wget ftp://ftp.ensembl.org/pub/release-110/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz

# Decompress
gunzip Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz

# Optional: rename to something simple
mv Homo_sapiens.GRCh38.dna.primary_assembly.fa GRCh38.fa

# Index for samtools / bcftools / CrossMap / Picard
samtools faidx GRCh38.fa
```

Download chainfile needed for liftover

```{bash}
wget http://hgdownload.soe.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz
```

Liftover Step\
When running from Initial directory `H3_Imputation.`

You can cp the files (`GRCh38, hg19TOHg38.over.chain.gz`) from the `reference` directory (after download) into your working `Liftover_GRCh38` directory.

```{bash}
cp /home/tafum/reference/GRCh38 /home/tafum/H3_Imputation/Lifover_GRCh38
cp /home/tafum/reference/hg19TOHg38.over.chain.gz /home/tafum/H3_Imputation/Liftover_GRCh38
```

#You should run this from `~/H3_imputation/Liftover_GRCh38`:

```{bash}
conda activate crossmap_env

CrossMap vcf \
  hg19ToHg38.over.chain.gz \
  H3_QC8_ibdClean_GRCh37.vcf.gz \
  GRCh38/GRCh38.fa \
  H3_QC8_ibdClean_GRCh38_lifted.vcf

```

```{bash}
# Sort by chrom+pos and write compressed VCF
bcftools sort \
  -Oz \
  -o H3_QC8_ibdClean_GRCh38_lifted.sorted.vcf.gz \
  H3_QC8_ibdClean_GRCh38_lifted.vcf
  
# Index the sorted VCF
tabix -p vcf H3_QC8_ibdClean_GRCh38_lifted.sorted.vcf.gz
```

### Convert sorted GRCh38 VCF → PLINK

```{bash}
plink --vcf H3_QC8_ibdClean_GRCh38_lifted.sorted.vcf.gz \
      --double-id \
      --make-bed \
      --out H3_QC8_ibdClean_GRCh38_tmp
```

### Restore original FAM and finalize GRCh38 dataset

```{bash}
cp H3_QC8_ibdClean.fam H3_QC8_ibdClean_GRCh38_tmp.fam

mv H3_QC8_ibdClean_GRCh38_tmp.bed H3_QC8_ibdClean_GRCh38.bed
mv H3_QC8_ibdClean_GRCh38_tmp.bim H3_QC8_ibdClean_GRCh38.bim
mv H3_QC8_ibdClean_GRCh38_tmp.fam H3_QC8_ibdClean_GRCh38.fam
```

#### Append QC to tracker

Run this from: `~/H3_imputation`

```{bash}
cd ~/H3_imputation

plink --bfile Liftover_GRCh38/H3_QC8_ibdClean_GRCh38 \
      --missing \
      --out Liftover_GRCh38/H3_QC8_ibdClean_GRCh38
      
Rscript qc_tracker.R Liftover_GRCh38/H3_QC8_ibdClean_GRCh38 \
  "Step 18: Liftover from GRCh37 to GRCh38 for imputation" \
  QC_progress_tracker.csv \
  --other "Started from H3_QC8_ibdClean (GRCh37); CrossMap vcf with hg19ToHg38.over.chain.gz and GRCh38.fa; 1,845,728 input variants, 885 failed to map; final PLINK dataset: H3_QC8_ibdClean_GRCh38.{bed,bim,fam}"

```

### Split GRCh38 dataset by chromosome

Move to your working directory:

```{bash}
cd ~ H3_Imputation/Liftover_GRCh38
```

#### Convert each chromosome to sorted VCF

Run this loop

```{bash}
for CHR in {1..22}; do
    plink --bfile chr${CHR}_GRCh38 \
          --recode vcf \
          --out chr${CHR}.raw

    # compress & sort & index
    bcftools sort -Oz -o chr${CHR}.vcf.gz chr${CHR}.raw.vcf
    tabix -p vcf chr${CHR}.vcf.gz

    # clean
    rm chr${CHR}.raw.vcf
done
```

#### Prepare imputation folder structure

```{bash}
mkdir -p ~/H3_imputation/imputation_input
mv chr*.vcf.gz* ~/H3_imputation/imputation_input/
```

Append to QC_tracker.csv (Do this from `~H3_Imputation`)

```{bash}
Rscript qc_tracker.R Liftover_GRCh38/H3_QC8_ibdClean_GRCh38   "Step 19: Prepared GRCh38 per-chromosome imputation dataset"   QC_progress_tracker.csv   --other "Split into chr1-22; converted to VCF; sorted bgzipped; indexed; ready for H3ABioNet chipimputation."
```

### Run H3ABioNet imputation

Make sure Nextflow and Singularity are on the HPC being used. If not you can create the environment using the following. Run:

```{bash}
cd ~
mkdir -p ~/bin
cd ~/bin

wget -qO nextflow https://github.com/nextflow-io/nextflow/releases/download/v24.04.3/nextflow
chmod +x nextflow
```

Now add `~/bin` to your PATH.

Edit your `.bashrc`:

```{bash}
echo 'export PATH="$HOME/bin:$PATH"' >> ~/.bashrc
source ~/.bashrc
```

Check installation

```{bash}
nextflow -v
```

You should see something like this: `nextflow version 24.04.3`

Liftover New Steps

Sort out lifover VCF

```{bash}
# If it's already gzipped and indexed partially, ignore the old index
# and just sort the VCF:

bcftools sort \
  -Oz -o H3_QC8_ibdClean_GRCh38_lifted_sorted.vcf.gz \
  H3_QC8_ibdClean_GRCh38_lifted.vcf.gz

tabix -p vcf H3_QC8_ibdClean_GRCh38_lifted_sorted.vcf.gz
```

#### Remove ambiguous SNPs

```{bash}
bcftools view -e '(REF="A" && ALT="T") || (REF="T" && ALT="A") || (REF="C" && ALT="G") || (REF="G" && ALT="C")' \
  -Oz -o H3_QC8_GRCh38_noAmbig.vcf.gz \
  H3_QC8_ibdClean_GRCh38_lifted_sorted.vcf.gz

tabix -p vcf H3_QC8_GRCh38_noAmbig.vcf.gz
```

#### **Fix alleles vs GRCh38 FASTA** (this is where strand flips get corrected):

```{bash}
bcftools +fixref H3_QC8_GRCh38_noAmbig.vcf.gz \
  -Oz -o H3_QC8_GRCh38_fixref.vcf.gz \
  -- -f /home/tafum/H3_imputation/Liftover_GRCh38/GRCh38/GRCh38.fa -m flip -m swap
```

#### Remove Indels and Multiallelic SNPs

```{bash}
bcftools view \
  -m2 -M2 -v snps \
  -Oz -o H3_QC8_GRCh38_fixref_snps.vcf.gz \
  H3_QC8_GRCh38_fixref.vcf.gz

tabix -p vcf H3_QC8_GRCh38_fixref_snps.vcf.gz
```

Set Base

```{bash}
BASE=H3_QC8_GRCh38_fixref_snps.vcf.gz
echo $BASE
```

Create chromosome rename map

```{bash}
cat > chr_rename.txt <<EOF
1   chr1
2   chr2
3   chr3
4   chr4
5   chr5
6   chr6
7   chr7
8   chr8
9   chr9
10  chr10
11  chr11
12  chr12
13  chr13
14  chr14
15  chr15
16  chr16
17  chr17
18  chr18
19  chr19
20  chr20
21  chr21
22  chr22
X   chrX
Y   chrY
EOF
```

Check the file

```{bash}
cat chr_rename.txt
```

Make sure it really is `old new` (e.g. `1   chr1`), not reversed.

## Run `--rename-chrs` *properly* on the combined VCF

Now run this **exact** command (don’t change order, don’t omit `$BASE`):

```{bash}
bcftools annotate \
  --rename-chrs chr_rename.txt \
  -Oz -o H3_QC8_GRCh38_fixref_snps_chr.vcf.gz \
  $BASE
```

If it works, it will be silent (no usage text). Now index:

```{bash}
tabix -p vcf H3_QC8_GRCh38_fixref_snps_chr.vcf.gz
```

Now **diagnose that new file**:

```{bash}
bcftools view -h H3_QC8_GRCh38_fixref_snps_chr.vcf.gz | grep '^##contig' | head
bcftools query -f '%CHROM\n' H3_QC8_GRCh38_fixref_snps_chr.vcf.gz | sort -u
```

You should now see:

-   Header lines like: `##contig=<ID=chr1,length=248956422,assembly=GRCh38.fa>`

-   And CHROM values: `chr1 chr2 … chr22 chrX`

#### Split chromosomes into Individual Chromosomes

``` {FINAL="H3_QC8_GRCh38_fixref_snps_chr.vcf.gz"}

for chr in {1..22} X; do
  echo "Extracting chr${chr} ..."
  bcftools view \
    -r chr${chr} \
    -Oz -o H3_QC8_GRCh38_fixref_snps_chr${chr}.vcf.gz \
    $FINAL

  tabix -p vcf H3_QC8_GRCh38_fixref_snps_chr${chr}.vcf.gz
done
```

#### Sanity Checks

Pick say chromosome 1 and inspect

```{bash}
bcftools query -f '%CHROM\n' H3_QC8_GRCh38_fixref_snps_chr1.vcf.gz | sort -u
```

Upload to Imputation Server

#### Download from Imputation Server

```{bash}
#!/bin/bash
set -e

PASSWORD="PUT_YOUR_PASSWORD_HERE"
LINK="https://impute.afrigen-d.org/get/nfHFNeBqt2eYrYrnAReI7ZQOMcajKnMVrdmEa6zD"

echo ""
echo "Downloading imputation results..."
curl -sL $LINK | bash -s "$PASSWORD"
echo ""
echo "All files downloaded successfully."
```

#### Unzip all chromosomes

```{bash}
cd ~/H3_imputation/final/imputated_data

for chr in {1..22}; do
  echo "Unzipping chr_${chr}.zip ..."
  unzip -o chr_${chr}.zip -d chr_${chr}
done
```

## Create the post-imputation QC + merge script

In `~/H3_imputation/final/imputated_data`, create this file:

```{bash}
nano post_impute_QC_and_merge.sh
```

Paste this in:

```{bash}
#!/bin/bash
set -euo pipefail

cd ~/H3_imputation/final/imputated_data

# ---------------- CONFIG ----------------
INFO_FIELD="R2"      # from VCF header: Estimated Imputation Accuracy (R-square)
INFO_THRESH=0.3      # you can increase to 0.5 or 0.8 if you want stricter QC
MAF_PLINK=0.01       # MAF threshold for PLINK QC
GENO_PLINK=0.02      # SNP missingness
MIND_PLINK=0.02      # sample missingness
# ----------------------------------------

echo "=== 1. Concatenate chunk VCFs per chromosome ==="

for chr in {1..22}; do
  echo ""
  echo ">>> Chromosome ${chr}: concatenating chunk VCFs"

  ls chr_${chr}/*.vcf.gz > chr_${chr}_vcf_list.txt

  bcftools concat -f chr_${chr}_vcf_list.txt -Oz -o imputed_chr${chr}.vcf.gz
  tabix -p vcf imputed_chr${chr}.vcf.gz
done

echo ""
echo "=== 2. VCF-level post-imputation QC (INFO/R2 + SNPs only) ==="

for chr in {1..22}; do
  IN_VCF="imputed_chr${chr}.vcf.gz"
  OUT_VCF="imputed_chr${chr}_filt.vcf.gz"

  echo ""
  echo ">>> Chromosome ${chr}: filtering ${IN_VCF} -> ${OUT_VCF}"

  bcftools view \
    -i "INFO/${INFO_FIELD}>=${INFO_THRESH}" \
    -m2 -M2 -v snps \
    -Oz -o "${OUT_VCF}" \
    "${IN_VCF}"

  tabix -p vcf "${OUT_VCF}"
done

echo ""
echo "=== 3. Convert filtered VCFs to PLINK per chromosome ==="

for chr in {1..22}; do
  VCF="imputed_chr${chr}_filt.vcf.gz"
  OUT="imputed_chr${chr}_plink"

  echo ""
  echo ">>> Chromosome ${chr}: converting ${VCF} -> ${OUT}.bed/bim/fam"

  plink \
    --vcf "${VCF}" \
    --double-id \
    --keep-allele-order \
    --make-bed \
    --out "${OUT}"
done

echo ""
echo "=== 4. Merge all chromosomes into one PLINK dataset ==="

BASE_PREFIX="imputed_chr1_plink"
MERGE_LIST="merge_list.txt"

rm -f "${MERGE_LIST}"

for chr in {2..22}; do
  echo "imputed_chr${chr}_plink.bed imputed_chr${chr}_plink.bim imputed_chr${chr}_plink.fam" >> "${MERGE_LIST}"
done

plink \
  --bfile "${BASE_PREFIX}" \
  --merge-list "${MERGE_LIST}" \
  --make-bed \
  --out imputed_all_chr_merged

echo ""
echo "=== 5. Final PLINK QC on merged imputed dataset ==="

plink \
  --bfile imputed_all_chr_merged \
  --geno ${GENO_PLINK} \
  --mind ${MIND_PLINK} \
  --maf ${MAF_PLINK} \
  --hwe 1e-6 midp \
  --make-bed \
  --out imputed_all_chr_QC

echo ""
echo "All done ✅"
echo "Final GWAS-ready dataset: imputed_all_chr_QC.{bed,bim,fam}"
```

Save and exit.

Then run:

```{bash}
chmod +x post_impute_QC_and_merge.sh
./post_impute_QC_and_merge.sh
```

## What you’ll end up with

-   Per-chromosome filtered VCFs: `imputed_chr${chr}_filt.vcf.gz`

-   Per-chromosome PLINK: `imputed_chr${chr}_plink.{bed,bim,fam}`

-   Merged, pre-QC PLINK: `imputed_all_chr_merged.{bed,bim,fam}`

**Final GWAS-ready dataset (after post-imputation QC):**\
👉 `imputed_all_chr_QC.{bed,bim,fam}`

#### Add unique IDs to each filtered per-chr VCFundefined

```{bash}
cd ~/H3_imputation/final/imputated_data  
for chr in {1..22}; do   echo "Setting IDs for chr${chr}..."   bcftools annotate \     --set-id '%CHROM:%POS:%REF:%ALT' \     -Oz -o imputed_chr${chr}_filt_id.vcf.gz \     imputed_chr${chr}_filt.vcf.gz    tabix -p vcf imputed_chr${chr}_filt_id.vcf.gz done
```

#### Quick sanity checks

```{bash}
bcftools query -f '%ID\t%CHROM\t%POS\t%REF\t%ALT\n' imputed_chr1_filt_id.vcf.gz | head # should look like: # 1:10583:A:G   1   10583   A   G
```

#### Concatenate all chromosomes into one whole genome VCF

```{bash}
Add unique IDs to each filtered per-chr VCFundefined

{bash}
cd ~/H3_imputation/final/imputated_data

for chr in {1..22}; do
  echo "Setting IDs for chr${chr}..."
  bcftools annotate \
    --set-id '%CHROM:%POS:%REF:%ALT' \
    -Oz -o imputed_chr${chr}_filt_id.vcf.gz \
    imputed_chr${chr}_filt.vcf.gz

  tabix -p vcf imputed_chr${chr}_filt_id.vcf.gz
done

Quick sanity checks

{bash}
bcftools query -f '%ID\t%CHROM\t%POS\t%REF\t%ALT\n' imputed_chr1_filt_id.vcf.gz | head
# should look like:
# 1:10583:A:G   1   10583   A   G

Concatenate all chromosomes into one whole genome VCF

{bash}
echo "Concatenating all chromosomes into one VCF..."
bcftools concat \
  imputed_chr1_filt_id.vcf.gz \
  imputed_chr2_filt_id.vcf.gz \
  imputed_chr3_filt_id.vcf.gz \
  imputed_chr4_filt_id.vcf.gz \
  imputed_chr5_filt_id.vcf.gz \
  imputed_chr6_filt_id.vcf.gz \
  imputed_chr7_filt_id.vcf.gz \
  imputed_chr8_filt_id.vcf.gz \
  imputed_chr9_filt_id.vcf.gz \
  imputed_chr10_filt_id.vcf.gz \
  imputed_chr11_filt_id.vcf.gz \
  imputed_chr12_filt_id.vcf.gz \
  imputed_chr13_filt_id.vcf.gz \
  imputed_chr14_filt_id.vcf.gz \
  imputed_chr15_filt_id.vcf.gz \
  imputed_chr16_filt_id.vcf.gz \
  imputed_chr17_filt_id.vcf.gz \
  imputed_chr18_filt_id.vcf.gz \
  imputed_chr19_filt_id.vcf.gz \
  imputed_chr20_filt_id.vcf.gz \
  imputed_chr21_filt_id.vcf.gz \
  imputed_chr22_filt_id.vcf.gz \
  -Oz -o imputed_all_filt_id.vcf.gz

tabix -p vcf imputed_all_filt_id.vcf.gz
```

Covert vcf to PLINK Files

```{bash}
plink \
  --vcf imputed_all_filt_id.vcf.gz \
  --double-id \
  --keep-allele-order \
  --make-bed \
  --out imputed_all_chr_raw
```

This will not contain sex and case/control stausts.

So use the original .fam file to append the sex and case/control status

```{bash}
awk '
  NR==FNR {
    # First pass: read original fam into maps
    sex[$2]   = $5   # $2 = IID, $5 = SEX
    pheno[$2] = $6   # $6 = PHENO
    next
  }
  {
    # Second pass: process imputed fam
    # FID/IID look like 250300001348_250300001348
    split($1, a, "_")
    core = a[1]            # 250300001348

    # Replace FID and IID with the core ID
    $1 = core
    $2 = core

    # Fill in SEX and PHENO from original fam
    $5 = sex[core]
    $6 = pheno[core]

    print
  }
' ~/Final_Analysis/H3_Africa_order_3220_original_annotation.fam \
  imputed_all_chr_raw.fam > imputed_all_chr_raw_annot.fam
```

Then replace the old `.fam` with the annotated one:

```{bash}
mv imputed_all_chr_raw_annot.fam imputed_all_chr_raw.fam
```

Quick sanity checks

```{bash}
head imputed_all_chr_raw.fam
```

You should see something like this:

`250300001348 250300001348 0 0 2 1250300001349 250300001349 0 0 1 1 250300001350 250300001350 0 0 2 1...`

#### Now run final QC on this annotated dataset

```{bash}
plink \
  --bfile imputed_all_chr_raw \
  --geno 0.02 \
  --mind 0.02 \
  --maf 0.01 \
  --hwe 1e-6 midp \
  --make-bed \
  --out imputed_all_chr_QC
```

Your **final GWAS-ready dataset** (with correct sex + case/control) is

`imputed_all_chr_QC.{bed,bim,fam}`

This is now the set you should use for all downstream GWAS (logistic regression, linear, etc.).

Builld a covariate file for the GWAS

```{bash}
# ============================
# Build covariate file WITH SEX
# ============================

pca_file   <- "/home/tafum/H3_imputation/PCA/H3_PCA.eigenvec"
pheno_file <- "/home/tafum/H3_imputation/final/GWAS/H3_pheno.phe.txt"
fam_file   <- "/home/tafum/Final_Analysis/H3_Africa_order_3220_original_annotation.fam"
out_file   <- "/home/tafum/H3_imputation/final/GWAS/H3_covariates_PC20.txt"

# -------------------------------
# 1. Read PCA eigenvectors
# -------------------------------
pca <- read.table(pca_file, header = FALSE, stringsAsFactors = FALSE)
n_pcs <- ncol(pca) - 2
pc_names <- paste0("PC", seq_len(n_pcs))
colnames(pca) <- c("FID", "IID", pc_names)

# -------------------------------
# 2. Read phenotype file (AGE + REGION)
# -------------------------------
pheno <- read.table(pheno_file, header = TRUE, stringsAsFactors = FALSE, sep = "\t")

# -------------------------------
# 3. Read the original fam to get SEX
# -------------------------------
fam <- read.table(fam_file, header = FALSE, stringsAsFactors = FALSE)
colnames(fam) <- c("FAM_FID", "IID", "FATHER", "MOTHER", "SEX", "PHENO")

# -------------------------------
# 4. Merge pheno + PCA
# -------------------------------
covar <- merge(
  pheno[, c("FID", "IID", "AGE", "REGION")],
  pca,
  by = c("FID", "IID"),
  all = FALSE
)

# -------------------------------
# 5. Add SEX from fam using IID
# -------------------------------
covar2 <- merge(
  covar,
  fam[, c("IID", "SEX")],
  by = "IID",
  all.x = TRUE
)

# -------------------------------
# 6. Keep AGE, REGION, SEX, and 20 PCs
# -------------------------------
max_pcs <- min(20, n_pcs)
pc_keep <- paste0("PC", 1:max_pcs)

covar_out <- covar2[, c("FID", "IID", "SEX", "AGE", "REGION", pc_keep)]

# -------------------------------
# 7. Write output
# -------------------------------
write.table(
  covar_out,
  file = out_file,
  quote = FALSE,
  sep = "\t",
  row.names = FALSE
)

cat("Covariate file written to:", out_file, "\n")
```

# Using H3AfricaBioNet Nextflow

Create a sample phenotype file that contains case-control status as `PHE`

```{bash}
cd /mnt/c/Users/theop/H3_Nextflow

# Header must have a column literally named "PHE" (case-sensitive)
printf "FID\tIID\tPHE\n" > sample.phe
awk -v OFS="\t" '{print $1, $2, $6}' H3_Africa_order_3220_original_annotation.fam >> sample.phe
```

Neftflow command

```{bash}
cd /mnt/c/Users/theop/H3_Nextflow/h3agwas

JAVA_HOME=/usr/lib/jvm/java-17-openjdk-amd64 NXF_VER=22.10.8 nextflow run qc/main.nf \
    -profile docker \
    --input_dir "/mnt/c/Users/theop/H3_Nextflow/" \
    --input_pat "H3_Africa_order_3220_original_annotation" \
    --output_dir "./qc_results" \
    -resume
```

Association analysis

```{bash}
cd /mnt/c/Users/theop/H3_Nextflow/h3agwas

# Clean everything
rm -rf work/ .nextflow/ ../assoc_results/

# THIS IS THE ONE THAT WORKS – give it the ORIGINAL files (not the QC’d ones)
JAVA_HOME=/usr/lib/jvm/java-17-openjdk-amd64 \
NXF_VER=22.10.8 \
nextflow run assoc/main.nf \
    -profile docker \
    --input_dir "/mnt/c/Users/theop/H3_Nextflow" \
    --input_pat "H3_Africa_order_3220_original_annotation" \
    --data "../sample.phe" \
    --pheno PHE \
    --output_dir "../assoc_results" \
    --output "H3_Africa_order_3220_original_annotation" \
    --assoc 1 \
    -with-timeline ../assoc_timeline.html \
    -with-report ../assoc_report.html
```

#### 
